---
title: "PhilaFire_PropensityScore"
author: "Ben Keel"
date: "2023-03-20"
output: html_document
---

## Set Up

```{r Setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,  message=FALSE, warning=FALSE)

library(boxr)
library(sf)
library(tidyverse)
library(knitr)
library(kableExtra)
library(tigris)
library(viridis)
library(dplyr)
library(tidycensus)
library(ggplot2)
library(lubridate)
library(janitor)
library(MatchIt)

options(scipen = 999)
mapTheme <- theme(plot.title =element_text(size=12),
                  plot.subtitle = element_text(size=8),
                  plot.caption = element_text(size = 6),
                  axis.line=element_blank(),
                  axis.text.x=element_blank(),
                  axis.text.y=element_blank(),
                  axis.ticks=element_blank(),
                  axis.title.x=element_blank(),
                  axis.title.y=element_blank(),
                  panel.background=element_blank(),
                  panel.border=element_blank(),
                  panel.grid.major=element_line(colour = 'transparent'),
                  panel.grid.minor=element_blank(),
                  legend.direction = "vertical", 
                  legend.position = "right",
                  plot.margin = margin(1, 1, 1, 1, 'cm'),
                  legend.key.height = unit(1, "cm"), legend.key.width = unit(0.2, "cm"))

#Color Palettes
palette3_sat <- c("#e19825","#d55816","#7b230b")
palette3_desat <- c("#B19C7D","#7F5F52","#262626")

palette4 <- c("#f1c82b","#e19825","#d55816","#7b230b")
palette4_desat <- c("#B19C7D","#B27D49","#7F5F52","#262626")

palette5_sat <- c("#f1c82b","#e19825","#d55816","#7b230b","#413028")
pallette5_desat <- c("#ead5b7","#d2b190","#b18e6f","#7f5f52","#413028")

palette6_sat <- c("white","#f1c82b","#e19825","#d55816","#7b230b","#413028")


palette7_cats <- c("#b9cfcf","#20b1ae","#e19825","#7b230b","#b47c49", "#3f3128", "#8f8172")

#Sources for Graphs
creditFire <- "Source: Philadelphia Fire Department"
creditOpen <- "Source: Open Data Philly"

g<-glimpse

source("https://raw.githubusercontent.com/urbanSpatial/Public-Policy-Analytics-Landing/master/functions.r")

crs <- "EPSG:4326"
PHL_crs <- 'ESRI:102411' #This measurement-based CRS projection is for the West Coast

```

# Process

1. Load in results, as we captured in EDA.

2. Add open data covariates (OPA, ACS, Transfer Values) to those results.

3. Check mean (0 to 1) or of results for each test group. 

# Tying the Sample Code to Our Own Data

<https://sejdemyr.github.io/r-tutorials/statistics/tutorial8.html>

## In the sample data:

Outcome variable of interest: standardized math score

Independent variable of interest: Catholic School y/n

Covariates:

-   race_white: Is the student white (1) or not (0)?

-   p5hmage: Mother's age

-   w3income: Family income

-   p5numpla: Number of places the student has lived for at least 4 months

-   w3momed_hsb: Is the mother's education level high-school or below (1) or some college or more (0)?

## In our data:

Outcome variable of interest: Vacancy / Permits / Transfers

Independent variable of interest: Fire y/n

-   Structure Fires Spatialized

    -   address (ID)

Covariates:

-   OPA Data (Tab Join)

    -   address (ID)
    
    -   building_code (categorical)
    
    -   total_livable_area (continuous)

    -   number_of_bedrooms (continuous)
    
    -   number_of_bathrooms (continuous)
    
    -   year_built (continuous)
    
    -   quality_grade (ordinal)
    
    -   owner occupied (t/f)
    
    -   assessment price/sqft (continuous)

-   ACS Data (Spatial Join)

    -   Tract Average Income (continuous)
    
    -   Tract average White racial makeup % (continuous)
    
-   Neighborhoods (Spatial Join)

    -   NHood Name
    
-   Transfers Data (optional)
    
    -   Sales price of 5 nearest buildings (continous) 

# Data Imports

## Fire Data

### Loading In

```{r Box Set Up, message=FALSE, warning=FALSE, include=FALSE}



```

### Cleaning

```{r Cleaning The Dataset,  message=FALSE, warning=FALSE,}
# Creating geometry for the fires
dat <- structureFire

dat <- dat %>% drop_na("Longitude", "Latitude") %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = "EPSG:4326")

# Cleaning the column names
dat <- clean_names(dat)

# Address string
dat <-
  dat %>% mutate(street_type = ifelse(street_type == 'AV', "AVE", street_type)) %>% 
  unite(address, c('address_number', 'street_prefix', 'street_name', 'street_type'), sep = " ", remove = FALSE, na.rm=TRUE)

# Extracting quarter
dat <- dat %>% mutate(quarter = floor_date(alarm_date, unit="quarter"))

# Reducing columns
dat <- dat %>%
  dplyr::select(address, quarter, property_use, incident_number, number_of_exposures, incident_type, building_status, fire_spread, no_spread, code_description, geometry, alarm_date, cad_nature_code_description,
                minor_damage, significant_damage, heavy_damage, extreme_damage)

# Removing duplicates
dat <- dat[!duplicated(dat$incident_number),]

dat_boolean <- dat%>%
  st_drop_geometry%>%
  mutate(fire = 1)%>%
  dplyr::select(address, fire)

taney <- dat %>% filter(grepl("N TANEY", address))

```

## ACS Philadelphia

### Variables

```{r Key and Variables}

census_api_key("3c9540be1434ac4b38e6e55d60e8ee95909f2254", overwrite = TRUE)

#Variables
acs_variable_list.2020 <- load_variables(2020, #year
                                         "acs5", #five year ACS estimates
                                         cache = TRUE)

acs_variable_list.2016 <- load_variables(2016, #year
                                         "acs5", #five year ACS estimates
                                         cache = TRUE)

```

### Loading



```{r ACS Data Loading, warning=FALSE, message=FALSE, results='hide'}

acs_vars <- c("B02001_001E", #Total Population
              "B02001_002E", #Total Population White Only
              "B19013_001E", #Median HH Income
              "B25002_001E", #Total Units
              "B25002_003E" #Total Vacant Units
              )

acsTractsPHL.2020 <- get_acs(geography = "tract",
                             year = 2020, 
                             variables = acs_vars, 
                             geometry = TRUE, 
                             state = "PA", 
                             county = "Philadelphia", 
                             output = "wide")%>%
  rename(Total_Pop = B02001_001E,
         Total_Pop_WhiteOnly = B02001_002E,
         Med_Income = B19013_001E,
         Total_Vacant = B25002_003E,
         Total_Units = B25002_001E)%>%
  mutate(Perc_White = Total_Pop_WhiteOnly/Total_Pop,
         Perc_Vacant = Total_Vacant*100/Total_Units)%>%
  dplyr::select(-B02001_001M, -B02001_002M, -B19013_001M, -Total_Pop_WhiteOnly)%>%
  st_set_crs("EPSG:4326")

```

## OPA Data

### Loading

```{r OPA Data Import}

# Reading the data
opa_dat <- read_csv("../Data/OpenDataPhilly-opa_properties_public.csv")

# Creating geometry for the properties
opa_dat <- opa_dat%>%
  drop_na(lat, lng)%>%
  st_as_sf(coords = c("lat", "lng"),
           crs = "EPSG:4326")

```

### Cleaning

```{r OPA Data Cleaning}

# Reducing columns
opa_ps <- opa_dat[!duplicated(opa_dat$location),] %>%
  dplyr::select(location, category_code, category_code_description, building_code, building_code_description, total_area, total_livable_area, market_value, mailing_street, number_of_bedrooms, number_of_bathrooms, number_stories, year_built, quality_grade)%>%
  filter(category_code == 1 | category_code == 2 | category_code == 3)%>%
  mutate(Price_Sqft = ifelse(is.na(total_livable_area) == FALSE & total_livable_area != 0, market_value / total_area, NA),
         quality_grade_mod = case_when(quality_grade == 1 ~ "E",
                                   quality_grade == 2 ~ "D",
                                   quality_grade == 3 ~ "C",
                                   quality_grade == 4 ~ "B",
                                   quality_grade == 5 ~ "A",
                                   quality_grade == 6 ~ "A+", 
                                   TRUE ~ quality_grade,
                                   ))%>%
  rename(address = location)%>%
  mutate(condo = ifelse(grepl("CONDO", building_code_description) == TRUE, TRUE, FALSE),
        owner_occ = ifelse(condo == FALSE & category_code_description != "MULTI FAMILY",
                           ifelse(address == mailing_street, TRUE, FALSE),
                           NA))%>%
  filter(Price_Sqft < 100000)%>%
  dplyr::select(-quality_grade)

opa_ps$quality_grade_mod <- factor(opa_ps$quality_grade_mod, order = TRUE,
                               levels = c("A+", "A", "A-", "B+", "B", "B-", "C+", "C", "C-", "D+", "D", "D-", "E+", "E", "E-"))


```

## Neighborhoods

### Loading In

```{r}

#https://github.com/azavea/geo-data

nhoods <- st_read("../Data/Neighborhoods_Philadelphia.geojson")%>%
  st_set_crs("EPSG:4326")

```


## Combining Data
The process for creating the set follows the steps in [this tutorial](https://simonejdemyr.com/r-tutorials/statistics/tutorial8.html). In summary, we create a rough model that estimates the likelihood that a building will have a fire, then match properties with similar propensity-for-fire scores and other variables using a matching algorithm. The initial contrast between the properties that had fires and those that didn't is here: 

```{r Tabular and Spatial Join }

#Fires
opa_fire <- opa_ps%>%
  left_join(dat_boolean, by="address")

opa_fire$fire <- as.factor(ifelse(is.na(opa_fire$fire), 0, 1))

#ACS and NHoods
opa_sf <- opa_fire%>%
  st_join(acsTractsPHL.2020)%>%
  st_join(nhoods)

```

# PreTests

```{r }

opa_sf_cov <- c('total_livable_area', 'number_of_bedrooms', 'Price_Sqft', 'Med_Income', 'owner_occ', 'Perc_White', 'quality_grade_mod', 'name')

opa_sf %>%
  filter(total_livable_area < 10000,
       Price_Sqft < 10000)%>%
  st_drop_geometry()%>%
  group_by(fire) %>%
  dplyr::select(one_of(opa_sf_cov)) %>%
  summarise_all(funs(mean(., na.rm = T)))

```

# Propensity Score Estimation

Here are the very rough variables for fire likelihood. We use predictors from the OPA dataset that monitor the quality of the home, the demographics of the area, include the current owner occupancy comparisons we previously engineered, and include the neighborhood name as a categorical variable to control for neighborhood effects. Again, this model could be vastly improved on its own, but the goal here is to simply get the overall sample distributions of these variables to be the same.

```{r, cache=TRUE, results='hide'}

m_ps2 <- glm(fire ~ total_livable_area + number_of_bedrooms + Price_Sqft + Med_Income + Perc_White + owner_occ + quality_grade_mod + name, 
            family = binomial(), data = opa_sf)

summary(m_ps2)

```

```{r}

prs_df <- data.frame(pr_score = predict(m_ps2, type = "response"),
                     fire = m_ps2$model$fire)
head(prs_df)

```


# Matching Algorithm

For the matching algorithm to work, we had to eliminate any observations that did not have the right data available. This cut down our count of fires from the original ~18,000 non-outliers we started with to around 12,000 incidents. Fortunately that's still a healthy sample.

See how the averages of the variables are much more aligned than before the modeling and matching process? With this set of 24,000 properties, we could move forward with analysis.

```{r, cache=TRUE}

opa_nomiss <- opa_sf %>%  # MatchIt does not allow missing values
  dplyr::select(address, fire, one_of(opa_sf_cov)) %>%
  na.omit()

mod_match <- matchit(fire ~ total_livable_area + number_of_bedrooms + Price_Sqft + Med_Income + Perc_White + owner_occ + quality_grade_mod + name,
                     method = "nearest", data = opa_nomiss)

```

```{r}

dta_m <- match.data(mod_match)

```

# Difference In Means

The means table is much closer after the matching algorithm.

```{r}

dta_m %>%
  st_drop_geometry()%>%
  group_by(fire) %>%
  dplyr::select(one_of(opa_sf_cov)) %>%
  summarize_all(funs(mean))

```

# Calculating Differences Between Prop-Matched Structures

## Loading In Data

```{r Loading Panel}

panel_opa <- read_csv("../Data/panel_OPA_Fire_OpenData.csv")

```

## Filter Panel By Addresses in Prop-Matched Set

```{r}

#assign fire time and incident number to one member of the subclass
dta_m_1 <- dta_m %>%
  filter(fire == 1)%>%
  left_join(st_drop_geometry(dplyr::select(dat, incident_number, address, quarter)))

#assign same data to other member of the subclass
dta_m_2 <- dta_m %>%
  filter(fire == 0)%>%
  left_join(st_drop_geometry(dplyr::select(dta_m_1, incident_number, quarter, subclass)), by="subclass")

dta_m_withTime <- rbind(dta_m_1, dta_m_2) #combine sets

panel_pm_4Y <- panel_opa%>%
  inner_join(dta_m_withTime, by="address")%>%
  mutate(mSinceFire = interval(quarter.y, quarter.x) %/% months(1),
         ySinceFire = mSinceFire / 12)%>%
  filter(mSinceFire >= -49,#Eliminate entries before fires (occurs because of incident_number group duplicates)
         mSinceFire < 49, #Eliminate entries after four years, as they are irrelevant (arbitrary)
         !(count > 0 & ySinceFire > 0))%>% #For addresses with multiple incidents, take out repeated fire observ's
  st_as_sf()

panel_pm_4Y$count <- as.factor(panel_pm_4Y$count)


```

# Graph Outcome Proximity to Fires

This is the essential table that the analysis provides. Fires are associates with increases in each of these outcomes.

```{r}

#Distribution of Outcomes, Measuring Time between Fire and Outcome
panel_pm_4Y%>%
  filter(year(quarter.y) < 2019 & year(quarter.y) > 2012)%>%
  st_drop_geometry%>%
  ungroup()%>%
  filter(!(n_Vacant == 0 & n_permits == 0 & n_transfers == 0))%>%
  dplyr::select(n_transfers, n_Vacant, n_permits, ySinceFire, fire)%>%
  gather(Variable, value, -ySinceFire, -fire)%>%
  filter(value > 0)%>%
    count(ySinceFire, fire, Variable)%>%
  ggplot(aes(x=ySinceFire, y=n, fill=fire))+
    geom_col(position="dodge", just = 0.5)+
      labs(title="Permits, Vacancies, and Sales By Time From Fire, Propensity-Matched Set",
       subtitle="Philadelphia County, 2013-2018, By Quarter")+
      geom_smooth(se=FALSE, aes(color = fire, group = paste(fire, (ySinceFire >= 0))), size = 1.5)+
  facet_wrap(~Variable) + 
  xlab("Years Since Fire")+
  ylab("Count of Outcomes")+
  scale_x_continuous(breaks = c(-4, -3, -2, -1, 0, 1, 2, 3, 4))+
    theme(panel.background = element_rect(fill = "#f3efe0"))+
  scale_fill_manual(values = c("#b18e6f", "#b9cfcf"))+
  scale_color_manual(values = c("#3f3128", "#20b1ae"))+
  geom_vline(xintercept = 0, color = "#d55816", size = 1, linetype = "dashed")


```

## Are investors buying fire-stricken properties at a higher rate than normal properties?

With our case study featuring two examples of investor-purchased homes, we were curious if this was a larger trend. In short, the transfer of properties between investors and homeowners is a minority of property transfers as a whole, but fires do affect sales in these categories in two different ways. 

```{r Importing sale transfer data}

transfers <- read_csv("../Data/OpenDataPhilly-transfers.csv")

#Select relevant fields and filter to fire data range
transfers_data <- transfers%>%
  dplyr:: select(objectid, recording_date, street_address, document_type, grantors, grantees)%>%
  filter(year(recording_date) > 2008,
         !is.na(street_address))%>%
  rename(address = street_address)%>%
  mutate(quarter = as_date(floor_date(recording_date, unit="quarter")))

```

We're defining investors as those who own above 5 properties in the OPA dataset and those that have 'LLC', 'LP', 'REO', 'Investments', 'Homes', 'Trust', 'Corp', 'Inc' in their name. From this criteria, we created a list of investors to find in the matched set.

```{r Creating an Investor List}

toMatch <- c('LLC', 'LP', 'REO', 'Investments', 'Homes', 'Trust', 'Corp', 'Inc')

investors <- unique(grep(paste(toMatch,collapse="|"), 
                        transfers_data$grantees, value=TRUE))

#Check for owners with 5+ properties
investors_opa <- opa_dat%>%
  st_drop_geometry()%>%
  dplyr::select(owner_1)%>%
  group_by(owner_1)%>%
  filter(n()>5)

investors_list_opa <- unique(investors_opa)

transfers_data <- transfers_data%>%
  mutate(investorBuy = ifelse(grantees %in% investors | grantees %in% investors_list_opa, TRUE, FALSE),
         investorSell = ifelse(grantors %in% investors | grantors %in% investors_list_opa, TRUE, FALSE))%>%
  mutate(homeownerToInvestor = ifelse(investorBuy == TRUE & investorSell== FALSE, TRUE, FALSE),
         investorToHomeowner = ifelse(investorSell == TRUE & investorBuy== FALSE, TRUE, FALSE))

#join transfers data to property by address and quarter

panel_pm_4Y_sales <- panel_pm_4Y%>%
  rename(quarter = quarter.x)%>%
  left_join(dplyr::select(transfers_data, address, quarter, homeownerToInvestor, investorToHomeowner), by=c("address", "quarter"))


```

### Homeowner to investor, investor to homeowner

Here’s the same type of graph as the property sales, but counting sales that we classify as “homeowner-to-investor” based on their name and how many properties they own. The rate of homeowner-to-investor sales is steadily increasing over time, but fires are associated with a nearly 100% increase in the first year, with the trend maintaining over time.

```{r homeowners to investors}

panel_pm_4Y_sales%>%
  filter(year(quarter.y) < 2019 & year(quarter.y) > 2012)%>%
  st_drop_geometry%>%
  ungroup()%>%
  filter(!(n_transfers == 0))%>%
  dplyr::select(ySinceFire, fire, homeownerToInvestor)%>%
  gather(Variable, value, -ySinceFire, -fire)%>%
  filter(value == TRUE)%>%
    count(ySinceFire, fire, Variable)%>%
  ggplot(aes(x=ySinceFire, y=n, fill=fire))+
    geom_col(position="dodge", just = 0.5)+
      labs(title="Property Transfers: Sales from Homeowners To Investors by Time to Fire",
       subtitle="Philadelphia County, Fires in 2013-2018, Sales in 2009-2022, By Quarter")+
      geom_smooth(se=FALSE, aes(color = fire, group = paste(fire, (ySinceFire >= 0))), size = 1.5)+
  facet_wrap(~Variable) + 
  xlab("Years Since Fire")+
  ylab("Count of Outcomes")+
  scale_x_continuous(breaks = c(-4, -3, -2, -1, 0, 1, 2, 3, 4))+
    theme(panel.background = element_rect(fill = "#f3efe0"))+
  scale_fill_manual(values = c("#b18e6f", "#b9cfcf"))+
  scale_color_manual(values = c("#3f3128", "#20b1ae"))+
  geom_vline(xintercept = 0, color = "#d55816", size = 1, linetype = "dashed")

```

Conversely, if we look at investor-to-homeowner sales, there’s a lag time where investor-owned properties with fires are sold at a lower rate, but then investor sales pick up after six months. This may indicate a flip-sale trend, with a purchase and renovation period in the couple quarters, followed by increased sales from the investors. 

```{r investors to homeowners}

panel_pm_4Y_sales%>%
  filter(year(quarter.y) < 2019 & year(quarter.y) > 2012)%>%
  st_drop_geometry%>%
  ungroup()%>%
  filter(!(n_transfers == 0))%>%
  dplyr::select(ySinceFire, fire, investorToHomeowner)%>%
  gather(Variable, value, -ySinceFire, -fire)%>%
  filter(value == TRUE)%>%
    count(ySinceFire, fire, Variable)%>%
  ggplot(aes(x=ySinceFire, y=n, fill=fire))+
    geom_col(position="dodge", just = 0.5)+
      labs(title="Property Transfers: Sales from Investors To Homeowners by Time to Fire",
       subtitle="Philadelphia County, Fires in 2013-2018, Sales in 2009-2022, By Quarter")+
      geom_smooth(se=FALSE, aes(color = fire, group = paste(fire, (ySinceFire >= 0))), size = 1.5)+
  facet_wrap(~Variable) + 
  xlab("Years Since Fire")+
  ylab("Count of Outcomes")+
  scale_x_continuous(breaks = c(-4, -3, -2, -1, 0, 1, 2, 3, 4))+
    theme(panel.background = element_rect(fill = "#f3efe0"))+
  scale_fill_manual(values = c("#b18e6f", "#b9cfcf"))+
  scale_color_manual(values = c("#3f3128", "#20b1ae"))+
  geom_vline(xintercept = 0, color = "#d55816", size = 1, linetype = "dashed")

```





